A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-11-07 imaging_demo/ms.tex
Minor changes throughout, update figures
From 1e9f59c (review):
=====================

Update MOFF and FX flowcharts, F-engine

In ms.tex,

(*) Minor changes throughout

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed layouts. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are being or were developed around the world based on this new 
paradigm, the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), including the Murchison 
Widefield Array (MWA;\citealt{tin13,bow13}), the Precision Array for Probing 
the Epoch of Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray 
(LOFAR;\citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment 
(CHIME,\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and 
the low frequency Square Kilometer Array (SKA1-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\sim\Ngrid\log_2\Ngrid$, where $\sim\Ngrid$ is the number of grid points in the 
Fourier transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often 
referred to as a ``direct imaging'' correlator because it eliminates the 
intermediate cross-correlation data products of the FX and XF correlators, but 
instead directly forms images from the electric field measurements.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes visibility based imaging using the software holography technique and a 
simulator for generating electric fields from a sky model.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and gridded. The gridded 
    electric fields $E_\textrm{g}(f)$ from each time series are imaged to produce 
    images $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to 
    obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $V_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix 
notation). They are calibrated and time-averaged to obtain 
$\langle V_\textrm{p}(f)\rangle$ which are then gridded and imaged to obtain the 
image $I(f)$. The $I(f)$ obtained from both techniques are theoretically 
identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $V_\textrm{p}(f)$ which are calibrated and time-averaged. Then 
    they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which 
    are then Fourier transformed to obtain the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Antenna-to-Grid Mapping}
\par\medskip
\noindent A grid is generated on the coordinate system in which antenna 
locations are specified with a grid spacing. The grid spacing can be controlled 
by the user. By default, it is set to be at $\lambda/2$ even at the 
lowest wavelength to ensure there is no aliasing even from regions of the sky 
far away from the field of view. The number of locations on the grid is 
restricted to be a power of 2 for efficient use of FFT.

@@

The antenna module is a fundamental building block upon which all the other 
modules are built. There is one antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectra.

@@

The critical component of the interferometer module is the X-engine. This is 
essentially a software analog of hardware correlators of real telescope 
systems. The X-engine can be toggled between two states of operation, 
namely, the FX and XF modes. The FX mode obtains the electric field spectra, 
$E(f)$ from the individual antenna modules inside this module and multiplies 
the two to obtain visibility spectra, $V(f)$. On the other hand, the XF mode 
cross-correlates the electric field time-series from its Antenna modules
to obtain the visibilities as a function of lags, $V_t(t)$, which is then Fourier 
transformed to obtain $V(f)$. Both modules can operate on dual-polarizations to
obtain all four cross-polarizations.

The other attributes (not shown in figure) are the flags applicable for each
cross-polarization for the current data stream. Similar to the antenna module,
it has an update function that can update the visibilities $V_t(t)$ or $V(f)$
directly rather than through the electric fields of its component antennas. 
This functionality is to allow EPIC to operate while attached to the backend 
of traditional correlator systems. This feature is not utilized for purposes 
of this paper.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format 
from 255 antennas within roughly a diameter of 100~m. The data is centered at a 
frequency of 74.03~MHz, with a sample rate (equal to the bandwidth) of 100~kHz 
with 512 complex time samples per antenna in a A/D writeout timescale of 
5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There are 
391 such writeouts (or time stamps; each contains 512 time samples at 100~kHz 
sampling ) yielding a total duration of 2~s.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over $\approx$~20~ms (four writeouts) of data 
and the inner $\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown 
in direction cosine coordinates -- $l$ along the horizontal axis and $m$ along 
the vertical axis. The flux scale is arbitrary. Even in this proof-of-concept 
demonstration, we see Cyg~A and Cas~A prominently as annotated, thus validating 
the functionality of the EPIC package.

@@

We elaborate on the data rates required with the MOFF and visibility based
algorithms. This is particularly relevant in the context of radio transient 
detection.

@@

It is seen to have significant savings in data throughput relative to a 
X-engine based pipeline. A unique advantage is the instantaneous availability 
of calibrated time-domain images at no extra cost. Hence, it is a compelling 
candidate for time-domain radio astronomy, e.g. search for and monitoring of 
transients. Potentially, it could allow on-chip processing thus lowering even 
further the relatively low I/O bandwidth shown in table~\ref{tab:data-rates} 
except when a transient event is detected. Transient detection pipelines at 
the backend of EPIC can be fine-tuned to target fast transients such as the 
Fast Radio Bursts \citep[FRB;][]{tho13} on millisecond timescales at GHz 
frequencies or slow transients from planetary and exoplanetary origins at 
frequencies around 100~MHz.

>>>>

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed layouts. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are being or were developed around the world based on this new 
paradigm, including the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), the Murchison Widefield Array 
(MWA;\citealt{tin13,bow13}), the Precision Array for Probing the Epoch of 
Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray 
(LOFAR;\citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment 
(CHIME,\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and 
the low frequency Square Kilometer Array (SKA1-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\sim\Ngrid\log_2\Ngrid$, where $\Ngrid$ is the number of grid points in the 
Fourier transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often 
referred to as a ``direct imaging'' correlator because it eliminates the 
intermediate cross-correlation data products of the FX and XF correlators, but 
instead directly forms images from the electric field measurements.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{The E-field Parallel Imaging Correlator (EPIC) package can be 
accessed at https://github.com/nithyanandan/EPIC} now publicly available. Besides 
implementing the MOFF imaging algorithm it also includes visibility based imaging 
using the software holography technique and a simulator for generating electric 
fields from a sky model.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$. They are calibrated 
    and gridded. The gridded electric fields $\widetilde{E}_\textrm{g}(f)$ from 
    each time series are imaged to produce images 
    $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to obtain the 
    final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $\widetilde{V}_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ 
in matrix notation). They are calibrated and time-averaged to obtain 
$\langle \widetilde{V}_\textrm{p}(f)\rangle$ which are then gridded and imaged to 
obtain the image $I(f)$. The $I(f)$ obtained from both techniques are 
theoretically identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $\widetilde{V}_\textrm{p}(f)$ which are calibrated and 
    time-averaged. Then they are gridded to obtain the gridded visibilities 
    $\widetilde{V}_\textrm{g}(f)$ which are then Fourier transformed to obtain 
    the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\par\medskip
\noindent {\bf Antenna-to-Grid Mapping}
\par\medskip
\noindent A grid is generated on the coordinate system in which antenna 
locations are specified with a grid spacing. The grid spacing can be controlled 
by the user. By default, it is set to be $\le\lambda/2$ even at the 
lowest wavelength to ensure there is no aliasing even from regions of the sky 
far away from the field of view. The number of locations on the grid is 
restricted to be a power of 2 for efficient use of FFT.

@@

The antenna module is a fundamental building block upon which all the other 
modules are built. There is one antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$\widetilde{E}(f)$ for both polarizations. The most important function inside 
this module is the F-engine that Fourier transforms time-series electric field 
data into spectra.

@@

The critical component of the interferometer module is the X-engine. This is 
essentially a software analog of hardware correlators of real telescope 
systems. The X-engine can be toggled between two states of operation, 
namely, the FX and XF modes. The FX mode obtains the electric field spectra, 
$\widetilde{E}(f)$ from the individual antenna modules inside this module and 
multiplies the two to obtain visibility spectra, $\widetilde{V}(f)$. On the 
other hand, the XF mode cross-correlates the electric field time-series from 
its Antenna modules to obtain the visibilities as a function of lags, 
$\widetilde{V}_t(t)$, which is then Fourier transformed to obtain 
$\widetilde{V}(f)$. Both modules can operate on dual-polarizations to obtain 
all four cross-polarizations.

The other attributes (not shown in figure) are the flags applicable for each
cross-polarization for the current data stream. Similar to the antenna module,
it has an update function that can update the visibilities $\widetilde{V}_t(t)$ 
or $\widetilde{V}(f)$ directly rather than through the electric fields of its 
component antennas. This functionality is to allow EPIC to operate while 
attached to the backend of traditional correlator systems. This feature is not 
utilized for purposes of this paper.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format 
from 255 antennas within roughly a diameter of 100~m. The data is centered at a 
frequency of 74.03~MHz, with a sample rate (equal to the bandwidth) of 100~kHz 
with 512 complex time samples per antenna in a A/D writeout timescale of 
5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There are 
391 such writeouts (or time stamps; each contains 512 time samples at 100~kHz 
sampling) yielding a total duration of 2~s.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over $\approx$~20~ms (four writeouts) of data 
and the inner $\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown 
in direction cosine coordinates -- $l$ along the horizontal axis and $m$ along 
the vertical axis. The flux scale is arbitrary. Even in this proof-of-concept 
demonstration, we see Cyg~A and Cas~A prominently as annotated, thus validating 
the functionality of EPIC.

@@

We elaborate on the I/O data rates required with the MOFF and visibility based
algorithms. This is particularly relevant in the context of radio transient 
detection.

@@

It is seen to have significant savings in data throughput relative to a 
X-engine based pipeline. A unique advantage is the instantaneous availability 
of calibrated time-domain images at no extra cost. Hence, it is a compelling 
candidate for time-domain radio astronomy, e.g. search for and monitoring of 
transients. Potentially, it could allow on-chip processing thus lowering even 
further the already relatively low I/O bandwidth shown in 
table~\ref{tab:data-rates} except when a transient event is detected. Transient 
detection pipelines at the backend of EPIC can be fine-tuned to target fast 
transients such as the Fast Radio Bursts \citep[FRB;][]{tho13} on millisecond 
timescales at GHz frequencies or slow transients from planetary and exoplanetary 
origins at frequencies around 100~MHz.
